---
title: "Assignment 3: Data Exploration"
author: "Jasmine Papas"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}
getwd()
library(tidyverse)
library(lubridate)
Neonics<- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv",stringsAsFactors = TRUE)
Litter<- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv",stringsAsFactors = TRUE )
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer:We may be interested in some of the environmental effects that the neonicotinoids may have. We might also want to learn which insects it will harm and how much harm it will cause. Farmers might be interested on the effects it may have on other crops or on human health if consumed. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Studying the litter and debris that falls to the forest floor is important for knowing what nutrients and minerals will be released into the soil. The types and amounts of each mineral/nutrient will decide the chemistry of the soil and determine what organisms may thrive and which may die in it. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.Elevated PVC litter traps
 2.temporal sampling-ground traps sampled once per year
 3. measured and classified by weight

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
dim(Neonics)
# 4623, 30
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r}
sort(summary(Neonics$Effect), decreasing= TRUE)
```

> Answer: Mortality and Population are the most common likely because they are the most interesting to farmers and those using the insecticides. The Mortality and Population will tell users how effective the insecticide is and how likely it is to solve their problem. 


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r}
summary(Neonics$Species.Common.Name)
```

> Answer: Honey Bee, Parasitic Wasp, Buff Tailed Bumblebee, Carniolan Honey Bee, Bumble Bee, Italian Honey bee. All of these are pollinators and therefore are crucial to the survival of a garden or farm. ADD MORE HERE

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author.)
```

> Answer: The dataset is a factor class because the concentrations are being used to categorize data and are not being used as numerical data. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r, fig.height = 3, fig.width = 4}
ggplot(Neonics)+
  geom_freqpoly(aes(x= Publication.Year), bins= 50)

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
#install.packages("wesanderson")
#library(wesanderson)
ggplot(Neonics)+
  geom_freqpoly(aes(x= Publication.Year, color= Test.Location), bins= 50)
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test location is the lab. However, before 2010, Feild Natural was the most common test location. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}
ggplot(Neonics)+
  geom_bar(aes(x=Endpoint))
theme(axis.title.x = element_text(angle = 90, vjust = 0.5, hjust =1))
```

> Answer:NOEL (Non-observable effected levels) and LOEL (Lowest-observable effected levels) are the two most common end points. 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
class(Litter$collectDate)
Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d") 
class(Litter$collectDate)
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r} 
summary(Litter$plotID)
unique(Litter$plotID)
```

> Answer: 12 plots were sampled at the Niwot Ridge. The unique function will get rid of any duplicates in the data while the summary function will just add them up. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r,fig.height= 3, fig.width= 4}
ggplot(Litter)+
  geom_bar(aes(x=functionalGroup))
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}
ggplot(Litter)+
  geom_boxplot(aes(x=dryMass))+
  geom_violin(aes(x=dryMass, y=functionalGroup))

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:The boxplot is more effective because it can better display the significant number of outliers than the violin plot can. 

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles
