---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Jasmine Papas"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1 Loading data and packages
library(tidyverse)
library(agricolae)
library(here)
library(lubridate)
library(base)

here()
NTL_chem_phys<-read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), stringsAsFactors = TRUE)
class("sampledate")
NTL_chem_phys$sampledate <- mdy(NTL_chem_phys$sampledate)

#2 My Theme
hw6_theme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "bottom")
theme_set(hw6_theme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temp. of July changes with change in depth. 
Ha: Mean lake temp. of July does not change with changes in depth.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
july.ntl<- NTL_chem_phys%>%
  mutate(month= month(sampledate))%>%
  select(lakename, year4, daynum, depth, temperature_C, month)%>%
  filter(month==7) %>% 
  drop_na()

#5 
july.ntl.scatter<-ggplot(july.ntl, aes(x=depth, y=temperature_C))+
  geom_point()+
  labs(x= "Depth", y= "Temperature")+
  ggtitle("Temperature vs Depth for July") 
  
print(july.ntl.scatter)


```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: It suggests that temperature does in fact change when depth changes. As the Depth increases, the temperature decreases. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
july.ntl.regression<- lm(july.ntl$temperature_C ~ july.ntl$depth)
summary(july.ntl.regression)


```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: With every 1m imcrease in depth, you can expect a 1.707 degree decrease in temperature. At base depth (depth=0) temperature will be approximately 20 degrees Celsius. There is significant correlation between temperature and depth (less than 0.001). 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
AIC <- lm(data = july.ntl, temperature_C ~ year4 + daynum + depth)
step(AIC)

#10
july.multi.regression <- lm(data = subset(july.ntl), 
                   temperature_C ~ depth + daynum + year4)
summary(july.multi.regression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 
The AIC suggests that we should use year4, daynum, and depth as variables to predict temperature. The multiple regression outcome showed that there was a significant correlation (less than 0.001) between all 3 variables. This shows us that there could be more than one reason that the temperature is changing, not just depth. 


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
july.anova <- aov(data = july.ntl, temperature_C ~ lakename)
summary(july.anova)

july.a.regression<- lm(data = july.ntl, temperature_C ~ lakename)
summary(july.a.regression)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: There is a significant different in mean temperature among the lakes. The ANOVA model showed that there is a significant correlation of less than 0.001 between temperature and lake. 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
july.temp<- ggplot(july.ntl, aes(x= depth, y= temperature_C ,color= lakename))+
            geom_point(alpha= 0.5)+
            geom_smooth(method = "lm", se= FALSE)+
            ylim(0, 35)+
            ggtitle("Temperature vs. Depth")+
            labs(x= "Depth", y= "Temperature (C)")
  
print(july.temp)
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(july.anova)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: 
Ward Lake has the same mean temperature as Peter Lake. East Long Lake and Hummingbird Lake are significantly distinct from all the other lakes.  

 Anything greater than 0.05 is the same

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: In order to compare mean temperatures between Peter and Paul Lake, we might also use a one-sample t test. 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
july.cramward<- july.ntl %>% 
  filter(lakename== "Crampton Lake"| lakename== "Ward Lake")


july.ttest<- t.test(july.cramward$temperature_C~ july.cramward$lakename)
july.ttest
```

>Answer: The test says that the mean temperatures are not equal and the t test shows that the p-value is insignificant. This does NOT match my answer for part 16. 
