---
title: "Assignment 10: Data Scraping"
author: "Student Name"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r, message = FALSE}
#1
library(tidyverse)
library(rvest)
library(base)
library(dplyr)
library(purrr)
getwd()


```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2
webpage<- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022')

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "36.1000".

```{r scrape.the.data}
#3
water.system.name <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
  
PWSID <- webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
  
ownership <-webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()

max.withdrawals.mgd <- webpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4
df_scraped<- data.frame("Month"= c("Jan", "May", "Sept", "Feb", "Jun", "Oct", "Mar","Jul", "Nov", "Apr", "Aug", "Dec"),
                        "Year"= rep(2022,12),
                        "Water_System_Name"= as.factor(water.system.name),
                        "Ownership"= as.factor(ownership),
                        "PWSID"= as.factor(PWSID),
                        "Max_Withdrawals"= as.numeric(max.withdrawals.mgd)
                        )

df_scraped<- df_scraped %>% 
  mutate(Date = my(paste(Month,"-",Year)))

#5
withdrawals_2022<-df_scraped %>% 
  ggplot(aes(x= factor(Month, level= c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")), y= Max_Withdrawals))+
    geom_point()+
    geom_line(group=1)+
  labs(title= "Monthly Maximum Withdrawal in 2022", x= "Month")
  
plot(withdrawals_2022)
```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6.
the_PWSID<- "03-63-020"
the_year<- 2022

scrape.it <- function(the_PWSID, the_year){
  thewebsite<- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', the_PWSID, '&year=', the_year))
  
  #Scrape the data items
  water.system.name <- thewebsite %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
  
PWSID <- thewebsite %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
  
ownership <-thewebsite %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()

max.withdrawals.mgd <- thewebsite %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
  #Convert to a dataframe
  df_scraped<- data.frame("Month"= c("Jan", "May", "Sept", "Feb", "Jun", "Oct", "Mar","Jul", "Nov", "Apr", "Aug", "Dec"),
                        "Year"= rep(the_year, 12),
                        "Water_System_Name"= as.factor(water.system.name),
                        "Ownership"= as.factor(ownership),
                        "PWSID"= as.factor(PWSID),
                        "Max_Withdrawals"= as.numeric(max.withdrawals.mgd)
                        )

df_scraped<- df_scraped %>% 
  mutate( Date = my(paste(Month,"-",Year)))
  
return(df_scraped)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
df_Durham_2015 <- scrape.it("03-32-010", 2015)

withdrawals_d2015<-df_Durham_2015%>% 
  ggplot(aes(x= factor(Month, level= c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")), y= Max_Withdrawals))+
    geom_point(color="darkgreen")+
    geom_line(group=1, color= "black")
  labs(title= "Monthly Maximum Withdrawal in Durham 2015", x= "Month")
  
plot(withdrawals_d2015)
```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8
df_Asheville_2015 <- scrape.it("01-11-010", 2015)
withdrawals_a2015<-df_Asheville_2015 %>% 
  ggplot(aes(x= factor(Month, level= c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")), y= Max_Withdrawals))+
    geom_point()+
    geom_line(group=1)+
    labs(title= "Monthly Maximum Withdrawal in Asheville 2015", x= "Month")
  
plot(withdrawals_a2015)

#combining the two
withdrawals_2015<-ggplot()+
    geom_point(data= df_Asheville_2015, mapping= aes(x= factor(Month, level= c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")), y= Max_Withdrawals), color="blue")+
    geom_point(data= df_Durham_2015, mapping= aes(x= factor(Month, level= c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")), y= Max_Withdrawals),color= "red")+
  labs(title= "Max Withdrawl in 2015; Asheville vs. Durham", x= "Month")
  
plot(withdrawals_2015)
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 
the_years<- c(2010:2021)
the_PWSID_<-"01-11-010"
dfs_2021<- map2(the_PWSID_, the_years, scrape.it) 
df_2021 <- bind_rows(dfs_2021)



#plot
withdrawals_2010_2021<-df_2021%>% 
  ggplot(aes(x= Date, y= Max_Withdrawals))+
    geom_point(color="red")+
    geom_smooth(method="loess")+
  labs(title= "Max Daily Withdrawals from 2010 to 2021", x= "Date")
  
plot(withdrawals_2010_2021)

```

> Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
Yes, there was a downward trend from 2010-2015 and an upward trend from 2015-2021. 